Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines

### 2013-03-24 paper/paper.tex
Claim that church was worst performing was bogus. Cleaned up some grammar.

In this article, we develop a systematic approach to deriving the
feasible implementation of an abstract-machine-based analyzer.

@@

This paper is organized in two halfs: in the first, we start with a
quick review of the AAM approach to develop an analysis framework and
then apply our step-by-step optimization techniques in the simplified
setting of a core functional language.  This allows us to explicate
the optimizations with a minimal amount of inessential technical
overhead.  In the second half, we scale this approach up to an
analyzer for a realistic untyped, higher-order imperative language
with a number of interesting features and then measure
improvements across a suite of benchmarks.

At each step during the initial presentation and development, we evaluate the
implementation on a benchmark from Vardoulakis and
Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
of multiplication over addition on Church numerals.
For the step-by-step development, 
this benchmark is particularly informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with
  complicated environment and control structure arising from the use
  of higher-order functions to encode arithmetic, and
%
\item it proves to be the \emph{least} improved benchmark of the
  complete suite considered in section~\ref{sec:eval}, and thus it
  serves as a good sanity check and lower-bound for each of the
  optimization techniques considered.
\end{enumerate}

We start, in section~\ref{sec:aam}, by developing an abstract
interpreter according to the AAM approach.  Without further
abstraction, the analysis is exponential due to per-state store
variance and thus cannot analyze the example in a reasonable amount of
time.  In section~\ref{sec:baseline}, we perform a further abstraction
by widening the store.  The resulting analyzer sacrifices precision
for speed and is able to analyze the example in about 1 minute.  This
step is described by Van Horn and Might~\cite[\S
  3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make
even small examples feasible.  We therefore take the widened
interpreter as the baseline for our evaluation.

@@

The AAM approach, in essence, does the following: it takes a
machine-based view of computation and turns it into a \emph{finitary
  approximation} by bounding the size of the store.  With a limited
address space, the store must map addresses to \emph{sets} of values.
Store updates are interpreted as joins, and store dereferences are
interpreted by non-deterministic choice of an element from a set.  The
result of analyzing a program is a finite directed graph where nodes
in the graph are (abstract) machine states and edges denote machine
transitions between states.

@@

ISWIM is a family of programming languages parameterized by a set of
base values and operations.  To make things concrete, we consider a
member of the ISWIM family with integers, booleans, and a few
operations.
%
Figure~\ref{fig:syntax} defines the (abstract) syntax of ISWIM.  It
includes variables, literals (either integers, booleans, or
operations), $\lambda$-expressions for defining procedures, procedure
applications, and conditionals.  Expressions carry a label, $\mlab$,
which is drawn from an unspecified set and denotes the source location
of the expression; labels are used to disambiguate distinct, but
syntactically identical pieces of syntax.  We omit the label
annotation in contexts where it is irrelevant.

@@

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval   &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\maddr}\\
&&|& \kar{\mexp,\menv,\maddr}\\
&&|& \kif{\mexp,\mexp,\menv,\maddr}\\
\text{Environments} &\menv&\in& \mathit{Var} \rightharpoonup \mathit{Addr}\\
\text{Stores} &\msto&\in& \mathit{Addr} \rightharpoonup \mathcal{P}(\mathit{Value})
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains} and
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as the set of states reachable by the
reflexive, transitive closure of the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values, not
  single values,
\item continuations are heap-allocated, not stack-allocated,
\item there are ``contour values'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\mathit{push}$, $\mathit{bind}$, and $\interpdelta$.
\end{itemize}

\begin{figure}
\begin{gather*}
\begin{align*}
\eval(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep&
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep&
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
\multicolumn{3}{r@{}}{
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)}
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ where } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}_\mlab^\mcntr(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is really a function).

@@

\paragraph{Abstract interpretation} To characterize abstract interpretation, set the implicit
parameters just as above, but drop the $\maddr \not\in \msto$
condition.  This family of interpreters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when multiple values reside in a store location.

@@

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

The uniform $k$-CFA allocation strategy would make $\eval$ in figure
\ref{fig:aam} a computable abstraction of reachable states, but not an
efficient one. This is not the strategy that AAM, nor we, recommend. Through
this and the following section, we will explain a succession of approximations
to reach the baseline analysis.  We'll compare performance at each stage to
identify the criticality of each optimization. 
%
We ground this journey by first formulating the analysis in terms of a classic
fixed-point computation.

@@

Conceptually, the AAM approach calls for computing an analysis as a graph
exploration: (1) start with an initial state, and (2) compute the transitive
closure of the transition relation from that state.

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentially to a merely
daunting polynomial.

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance \emph{and} precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis. The
difference is that the store is not modified in-place, but updated
after all frontier states have been processed. This has implications
for the analysis' precision and determinism. Specifically, higher
precision, and is deterministic even if set iteration is not.

@@

@@

The final approximation we make to get to our point of departure is
store-allocating results of application sub-expressions. 
The AAM approach stops at the previous optimization. 
However, the {\tt fn} continuation stores a value, and this makes the space of
continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0CFA.
Having the space of continuations grow linearly with the size of the program 
will drop the overall complexity to cubic (as expected).

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. In our example language, we have only
unary functions, so application forms always have two positions: the
function and the argument. Thus for every application labeled $\mlab$,
we introduce two new labels, $\mlab^f$ and $\mlab^a$.
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

In $k$-CFA, the evaluation rules become:

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) transition elimination
and (2) pragmatic improvement.
%
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item store deltas;
 \item timestamped stores; and
 \item preallocated data structures.
\end{enumerate}

All but the first technique are precision preserving. We will discuss the design rationale
for the move.

@@

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces two-way branching, only to be ultimately joined back together in their respective
application positions. 
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.3]{fanout}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). 
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.3]{lazy}
\end{center}

>>>>

In this article, we develop a systematic approach to deriving a
feasible implementation of an abstract-machine-based analyzer.

@@

We start with a quick review of the AAM approach to develop an
analysis framework and then apply our step-by-step optimization
techniques in the simplified setting of a core functional language.
This allows us to explicate the optimizations with a minimal amount of
inessential technical overhead.  Following that, we scale this
approach up to an analyzer for a realistic untyped, higher-order
imperative language with a number of interesting features and then
measure improvements across a suite of benchmarks.

At each step during the initial presentation and development, we
evaluated the implementation on a set of benchmarks. The highlighted
benchmark in figure \ref{fig:churchtime} is from Vardoulakis and
Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
of multiplication over addition on Church numerals.  For the
step-by-step development, this benchmark is particularly informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with
  complicated environment and control structure arising from the use
  of higher-order functions to encode arithmetic, and
%
\item its improvement is about median in the benchmark suite
  considered in section~\ref{sec:eval}, and thus it serves as a good
  sanity check for each of the optimization techniques considered.
\end{enumerate}

We start, in section~\ref{sec:aam}, by developing an abstract
interpreter according to the AAM approach.  In the initial
abstraction, each state carries a store (what is called per-state
store variance). The space of stores is exponential in size; without
further abstraction, the analysis is exponential and thus cannot
analyze the example in a reasonable amount of time.  In
section~\ref{sec:baseline}, we perform a further abstraction by
widening the store.  The resulting analyzer sacrifices precision for
speed and is able to analyze the example in about 1 minute.  This step
is described by Van Horn and Might~\cite[\S
3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make even
small examples feasible.  We therefore take a widened interpreter as
the baseline for our evaluation.

@@

The AAM approach, in essence, does the following: it takes a
machine-based view of computation and turns it into a \emph{finitary
  approximation} by bounding the size of the store.  With a limited
address space, the store must map addresses to \emph{sets} of values.
Store updates are interpreted as joins, and store dereferences are
interpreted by non-deterministic choice of an element from a set.  The
result of analyzing a program is a finite directed graph where nodes
in the graph are machine states and edges denote machine
transitions between states.

@@

ISWIM is a family of programming languages parameterized by a set of
base values and operations.  To make things concrete, we consider a
member of the ISWIM family with integers, booleans, and a few
operations.
%
Figure~\ref{fig:syntax} defines the syntax of ISWIM.  It
includes variables, literals (either integers, booleans, or
operations), $\lambda$-expressions for defining procedures, procedure
applications, and conditionals.  Expressions carry a label, $\mlab$,
which is drawn from an unspecified set and denotes the source location
of the expression; labels are used to disambiguate distinct, but
syntactically identical pieces of syntax.  We omit the label
annotation in contexts where it is irrelevant.

@@

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\maddr}\\
&&|& \kar{\mexp,\menv,\maddr}\\
&&|& \kif{\mexp,\mexp,\menv,\maddr}\\
\text{Addresses} &\maddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains} and
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as its traces given by the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values, not
  single values,
\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\mathit{alloc}$, $\mathit{tick}$,
  $\mathit{push}$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

More generally, we can make stores map to a domain that forms a galois
connection with sets of values, so that $\interpdelta$ can give more
elaborate abstractions of base values (e.g., with the interval
abstraction, $\{0,[0,5]\} \equiv \{[0,5]\}$). We stick with sets for
simpler exposition.

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
&&\mcntr' =\tick(\mcntr) \\
&&\maddr  =\alloc(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\push_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
$\mcntr$, but they can be used to determinize the choice of fresh
addresses.  The resulting relation is non-deterministic in its choice
of addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is
really a function).

@@

\paragraph{Abstract interpretation} To characterize abstract
interpretation, set the implicit parameters just as above, but drop
the $\maddr \not\in \msto$ condition. $\interpdelta$ takes some care
to not make the analysis run forever; a simple instantiation is a flat
abstraction that returns an abstract top element $\mathit{Integer}$ as
the result of arithmetic operations.  This family of interpreters is
also non-deterministic in choices of addresses, but it is free to
choose addresses that are already in use.  Consequently, the machines
may be non-deterministic when multiple values reside in a store
location.

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\push_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\msto,\mval}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

The uniform $k$-CFA allocation strategy would make $\traces$ in figure
\ref{fig:aam} a computable abstraction of possible executions, but not an
efficient one. This is not the strategy that AAM, nor we, recommend. Through
this section, we will explain a succession of approximations
to reach the baseline analysis.  We'll compare performance at each stage to
identify the criticality of each choice. 
%
We ground this journey by first formulating the analysis in terms of a classic
fixed-point computation.

@@

Conceptually, the AAM approach calls for computing an analysis as a
graph exploration: (1) start with an initial state, and (2) compute
the transitive closure of the transition relation from that state. All
visited states are potentially reachable in the concrete, and all
paths through the graph are possible traces of execution.

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

The possible traces of execution tell us the most about a program, so
we take $\traces(\mexp)$ to be the (regular) set of paths through the
computed graph.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentiality to a merely
daunting polynomial.

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance \emph{and} precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis. The
difference is that the store is not modified in-place, but updated
after all frontier states have been processed. This has implications
for the analysis' precision and determinism. Specifically, higher
precision, and it is deterministic even if set iteration is not.

@@

For clarity, we will exhibit non-widened semantics unless a technique
specifically requires it.

@@

The final approximation we make to get to our point of departure is
store-allocating results of application sub-expressions. 
The AAM approach stops at the previous optimization. 
However, the {\tt fn} continuation stores a value, and this makes the space of
continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0-CFA.
Having the space of continuations grow linearly with the size of the program 
will drop the overall complexity to cubic (as expected).

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate addresses to store the values of each
subexpression of an application. The continuations have to store these
addresses to remember where to get argument values from, though
sometimes your allocation strategy may allow you to elide that
representation.

The new evaluation rules follow (where $\mcntr' = \tick(\mlab,\mcntr)$).
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) transition elimination
and (2) pragmatic improvement.
%
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
  \setcounter{enumi}{2}
 \item store deltas;
 \item timestamped stores; and
 \item preallocated data structures.
\end{enumerate}

All techniques are precision preserving under reasonable
assumptions. We will discuss these assumptions as the caveats arise.

@@

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces $n$-way branching, only to be ultimately joined back together in their respective
application positions. 
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.3]{fanout}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). 
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.3]{lazy}
\end{center}
